---
title: "Homework 4"
author: "Sonia Kangaju & Abbey Skinner"
date: "2026-03-27"
output: pdf_document
---

```{r, include = FALSE}
library(riskCommunicator)
library(ggplot2)
library(dplyr)
library(tidyr)
library(gt)
library(kableExtra)
library(gee)
```

```{r, include = FALSE}
framingham <- read.csv("~/Desktop/UNC/Spring 2026/Bios 767/Homeworks/bios767finalproject/framingham.csv")
framingham_wide <- read.csv("~/Desktop/UNC/Spring 2026/Bios 767/Homeworks/bios767finalproject/framingham_wide.csv")
```


# Mean Model Selection

**1. Candidate mean structures (15 points): Fit at least three candidate mean models to your data. These might include a saturated (profile) model, polynomial models (linear, quadratic), or spline models. For each model, report the number of parameters and the log-likelihood.**

```{r, include = FALSE}
# Pivot data to long form for model fitting
framingham_long = framingham_wide %>% pivot_longer(sysbp1:sysbp3,
                                                   names_to = "PERIOD", values_to = "SYSBP")
framingham_long$PERIOD = rep(1:3, nrow(framingham_wide)) %>% factor(levels = c(1,2,3))
framingham_long$YEAR = rep(c(0,6,12), nrow(framingham_wide))
framingham_long$SPLINE = pmax(framingham_long$YEAR - 6, 0)
framingham_long$SEX = (framingham_long$SEX == 2) %>% as.numeric()
```

As established in the previous assignment, we will use AR(1) as our working covariance model to compare mean models. First, we fit a saturated model, where the time period is treated as a categorical variable.

```{r, echo = FALSE}
# Saturated model with working AR(1) covariance
fit_sat = gee(SYSBP ~ PERIOD * SEX,
              data = framingham_long,
              id = RANDID,
              corstr = "AR-M", Mv = 1)
coef(summary(fit_sat))
```

Next, we fit a linear model, where time is treated as a continuous variable.

```{r, echo = FALSE}
# Linear model with working AR(1) covariance
fit_linear = gee(SYSBP ~ YEAR * SEX,
                 data = framingham_long,
                 id = RANDID,
                 corstr = "AR-M", Mv = 1)
coef(summary(fit_linear))
```

Finally, we fit a linear spline model, with a knot at year 6.

```{r, echo = FALSE}
# Spline model with working AR(1) covariance
fit_spline = gee(SYSBP ~ YEAR + SPLINE + SEX + YEAR:SEX + SPLINE:SEX,
                 data = framingham_long,
                 id = RANDID,
                 corstr = "AR-M", Mv = 1)
coef(summary(fit_spline))
```

Each model estimates a scale parameter and a correlation parameter that jointly define the AR(1) covariance structure. In addition to this, the saturated mean model consists of 6 parameters, the linear mean model consists of 4 parameters, and the spline mean model consists of 6 parameters. 

These GEE models make no distributional assumptions about the outcome, so there are no log-likelihoods to report.

**2. Model comparison (15 points): Compare your candidate mean models using appropriate criteria. Select a mean structure and justify your choice. If you conduct any hypothesis tests to compare nested models, explain your approach.**

The linear model is nested in the spline model, so we can compare them using a Wald test with $H_0: \beta_{SPLINE} = \beta_{SPLINE:SEX} = 0$. We can also write this as $Q'\beta = \boldsymbol{0}$ where $Q' = \left( \begin{smallmatrix}  0 & 0 & 1 & 0 & 0 & 0\\ 0 & 0 & 0 & 0 & 0 & 1 \end{smallmatrix} \right)$ and $\boldsymbol{\beta} = (\beta_0, \beta_1, \beta_2, \beta_3, \beta_4, \beta_5)^T$.

$T_W = (\boldsymbol{Q'}\widehat{\boldsymbol{\beta}})'[\boldsymbol{Q'}\widehat{Var}(\widehat{\boldsymbol{\beta}})Q]^{-1}(\boldsymbol{Q'}\widehat{\boldsymbol{\beta}})$ ~ $\chi^2_2$ under $H_0$. 

After running the test with 2 degrees of freedom, we get a test statistic of 4.44 and a p-value of 0.108. Since the p-value is $>0.05$, we fail to reject the null hypothesis and conclude that the spline model does not explain significantly more of the variation in SBP over time, when compared to the linear model.

```{r, echo = FALSE}
# Q below is Q', but labeled as Q for ease of coding
Q <- rbind(
  c(0, 0, 1, 0, 0, 0),
  c(0, 0, 0, 0, 0, 1)
)
beta.hat = fit_spline$coefficients
cov.robust = fit_spline$robust.variance

# Wald statistic
Qbeta <- Q %*% beta.hat
VQ    <- Q %*% cov.robust %*% t(Q)

Wald_stat <- as.numeric(t(Qbeta) %*% solve(VQ) %*% Qbeta)
df_wald   <- qr(Q)$rank
p_value   <- pchisq(Wald_stat, df = df_wald, lower.tail = FALSE)

wald_results <- data.frame(
  Statistic = Wald_stat,
  DF = df_wald,
  P_value = p_value
)

wald_results
```

From here on, we compare the linear model to the saturated model. Since the GEE model is not likelihood-based and the two remaining models are non-nested, we will compare their goodness of fit using QIC.

```{r}
QIC(fit_sat, fit_linear)
```

The saturated model and the linear model, which suggests that neither model fits the data better than the other. In light of this, we will err on the side of caution and choose the model that imposes fewer assumptions on the shapes of the mean SBP profiles. Therefore, we will choose the saturated model as our final mean model.


# Covariance Model Selection

**1. Candidate covariance structures (15 points): Using your selected mean structure, fit your model with at least three different covariance structures. For each covariance structure, report the number of covariance parameters and the log-likelihood.**

As noted before, we will use the saturated model as our mean structure. The three covariance structures that we will test are AR(1) (used in mean model selection), unstructured, and compound symmetry. Both the AR(1) and compound symmetry models estimate 2 covariance parameters, $\sigma^2$ and $\rho$. The unstructured covariance model estimates $n(n+1)/2 = 3(4)/2 = 6$ covariance parameters. The table below gives the AIC, BIC, and REML log-likelihood that each model yields.

```{r, include = FALSE}
# AR(1)
fit_ar1 <- gls(
  SYSBP ~ PERIOD*SEX,
  data = framingham_long,
  corr = corAR1(form = ~ as.numeric(as.character(PERIOD)) | RANDID),
  method = "REML"
)

# Unstructured
fit_un <- gls(
  SYSBP ~ PERIOD*SEX,
  data = framingham_long,
  corr = corSymm(form = ~ 1 | RANDID),
  weights = varIdent(form = ~ 1 | PERIOD),
  method = "REML"
)

# Compound Symmetry
fit_cs <- gls(
  SYSBP ~ PERIOD*SEX,
  data = framingham_long,
  corr = corCompSymm(form = ~ 1 | RANDID),
  method = "REML"
)

model_summary <- data.frame(
  Model = c("Compound Symmetry", "AR(1)", "Unstructured"),
  Cov_Params = c(2, 2, 6),
  AIC = c(AIC(fit_cs), AIC(fit_ar1), AIC(fit_un)),
  BIC = c(BIC(fit_cs), BIC(fit_ar1), BIC(fit_un)),
  LogLik = c(as.numeric(logLik(fit_cs)),
             as.numeric(logLik(fit_ar1)),
             as.numeric(logLik(fit_un)))
)
```


```{r, echo = FALSE}
model_summary %>%
  gt() %>%
  tab_header(
    title = "Model Comparison",
    subtitle = "Covariance Structure Selection"
  ) %>%
  cols_label(
    Model = "Covariance Structure",
    Cov_Params = "# Cov Params",
    AIC = "AIC",
    BIC = "BIC",
    LogLik = "Log-Likelihood"
  ) %>%
  fmt_number(
    columns = c(AIC, BIC, LogLik),
    decimals = 2
  )
```


**2. Model comparison (15 points): Compare your candidate covariance structures using appropriate criteria. Select a covariance structure and justify your choice. If any comparisons involve testing whether a variance component equals zero, discuss how the boundary issue affects your approach.**

Compound Symmetry and AR(1) are both nested within Unstructured covariance so we can compare these with REML LRT. To compare CS and AR(1) we need to use AIC since these are not nested. When comparing compound symmetry to unstructured, the REML LRT yields a test statistic of $226.98$ and a p-value of $<.0001$. The REML LRT for comparing AR(1) to unstructured yields a test statistics of $340.04$ and a p-value of $<.0001$. These tests tell us that in both cases, unstructured covariance is preferred over the simpiler structures. For completeness, we found that AIC for CS is $82027.30$ and for AR(1) the AIC is $82140.36$ which leads to us choosing compound symmetry over AR(1) since it has a smaller AIC. Based on all these criteria, we will choose unstructured covariance for our final model. Although it requires estimating more paramters, we found statistically significant evidence that estimating more parameters improves our model.

Note that using REML LRT is not valid with when we are testing that a variance component equals zero. This is becuase variances are bounded below by zero and when we test if a variance equals 0 this places the null hypothesis on the boundary of the parameter space. The LRT is only valid when the null value is on the interior of the parameter space, so this test is not valid in cases when a variance parameter equals 0. This does not arise in our case.

```{r, include = FALSE}
# compare CS and AR(1) to unstructured
anova(fit_cs, fit_un)
anova(fit_ar1, fit_un)
```


# Final Model

**1. Model specification (10 points): State your final model, including the mean structure and covariance structure. Explain why you chose these structures.**

Our final model uses the saturated model as the mean model and unstructured covariance for the covariance structure. We chose the saturated model since it imposes fewer assumptions than a linear model and we found that both the saturated model and the linear model fit the data similarly (when compared using QIC). We chose unstructured covariance since we found that it fits the data better than both compound symmetry and AR(1) when compared using REML LRT.

**2. Parameter estimates (15 points): Report the fixed effect estimates with standard errors and 95% confidence intervals. Report the covariance parameter estimates. Interpret the key parameters in the context of your research question.**

Find all estimates, standard errors, 95% confidence intervals, and p-values in the table below. The estimated covariance matrix is provided below the table. $\widehat{\beta_0} = 129.02$ represents the mean SBP for males at the first visit. $\widehat{\beta_1} = 5.55$ and $\widehat{\beta_2} = 10.25$ represent the change in SBP from baseline to visits 2 and 3 respectively. $\widehat{\beta_4} = 1.92$ represents the difference in SBP between males and females at baseline. All of these terms are significant indicating significant differences in SBP both over time and between males and females (evaluated separately). $\widehat{\beta_5} = -0.01$ and $\widehat{\beta_6} = -0.34$ represents the difference in SBP between males and females at later visits. Neither of these terms is significant indicating that the difference in SBP between males and females does not change over time

```{r, echo=FALSE}
# Fixed effects summary
beta_tab <- as.data.frame(summary(fit_un)$tTable)
beta_tab$Term <- rownames(beta_tab)
rownames(beta_tab) <- NULL

# Confidence intervals
beta_ci <- as.data.frame(intervals(fit_un, which = "coef")$coef)
beta_ci$Term <- rownames(beta_ci)
rownames(beta_ci) <- NULL

# Merge into one clean table
fixed_effects_table <- beta_tab %>%
  select(Term, Value, Std.Error, `p-value`) %>%
  left_join(beta_ci %>% select(Term, lower, upper), by = "Term") %>%
  rename(
    Estimate = Value,
    SE = Std.Error,
    p_value = `p-value`,
    CI_lower = lower,
    CI_upper = upper
  )

fixed_effects_table %>%
  gt() %>%
  tab_header(
    title = "Final Model",
    subtitle = "Saturated Mean Model, Unstructured Covariance"
  ) %>%
  cols_label(
    Estimate = "Estimate",
    SE = "Standard Error",
    p_value = "P-Value",
    CI_lower = "CI-Lower",
    CI_upper = "CI-Upper"
  ) %>%
  fmt_number(
    columns = c(Estimate, SE, p_value, CI_lower, CI_upper),
    decimals = 2
  )
```

```{r, echo=FALSE}
# Residual SD
sigma_hat <- fit_un$sigma

# Variance multipliers (relative to reference timepoint)
var_weights <- coef(fit_un$modelStruct$varStruct, unconstrained = FALSE)

# Correlation parameters
cor_params <- coef(fit_un$modelStruct$corStruct, unconstrained = FALSE)

# Variances
var_mult <- c(1, var_weights)^2
vars <- sigma_hat^2 * var_mult

# Correlation matrix
R_hat <- matrix(c(
  1,             cor_params[1], cor_params[2],
  cor_params[1], 1,             cor_params[3],
  cor_params[2], cor_params[3], 1
), nrow = 3, byrow = TRUE)

# Covariance matrix
D_half <- diag(sqrt(vars))
Sigma_hat <- D_half %*% R_hat %*% D_half

Sigma_hat
```



# Visualization and Interpretation

**1. Visualization (10 points): Create a figure showing the observed data (e.g., group means over time) overlaid with the fitted trajectories from your final model.**

**2. Summary (5 points): In 1–2 paragraphs, summarize your model-building process and key findings as you would for a collaborator.**
